---
title: "Age and Performance in Masters Swimming and Running"
subtitle: "A Simplified Example"
author: "R. De Veaux, A. Plantinga, and E. Upton"
date: "2024-06-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=60), tidy=TRUE, message = FALSE, warning = FALSE)

library(tidyverse) 
library(stringi)
library(lubridate)
library(broom)
library(janitor)
library(kableExtra)
library(gridExtra)
library(doParallel)
library(dplyr) 
library(progress)
library(purrr)
library(stacks)
library(stringr)
library(tidymodels)

```

## Introduction and data

This document illustrates the bootstrap modeling process.  For this simplified walk through, we focus on one event.  Note, the data for the entire analysis are available from the websites cited in our manuscript. For this working example, to ease reproducibility, we include the file `swim_1650.RDS` (data from the 1650Y swim).  

```{r}

## Load data 
swimming <- readRDS("swim_1650.RDS")

swimming<- swimming %>% 
  filter( Age %% 5 == 0, 29 < Age, Age < 81) %>%
  mutate(Female = ifelse(Sex == "M", 0, 1))

## Explore data
g <- ggplot(swimming, aes(x = Age, y = TimeSec, color = Sex, shape = Sex)) +
  geom_smooth(se = TRUE, method = 'loess', formula = 'y ~ x') +
  geom_point(alpha = .1) 
```

Define the response (ratio to age 35):

```{r}
#Calculating ratio (for response)
swimming_current_m <- swimming %>%
  group_by(Sex)%>%
  filter(Age == 35) %>%
  summarize(mean(TimeSec))

swimming_current_m <- as.matrix(swimming_current_m)

swimming <- swimming  %>%
  mutate(age35 = ifelse(Sex == 0, 
  as.numeric(swimming_current_m[1,2]), 
  as.numeric(swimming_current_m[2,2]))) %>%
  mutate(Ratio = TimeSec / age35)

```

Define the stacked model:

```{r}
do.one <- function(seed, swim,  nhidden, nepoch, ndeg, spdeg) {
  
  mAge <- mean(swim$Age)
  sAge <- sd(swim$Age) 
  mRatio <- mean(swim$Ratio) 
  sRatio <- sd(swim$Ratio)
  
  #standardize
  this_swim <- swim %>% 
    mutate(zAge = (Age - mAge)/sAge, 
           zAgeF = zAge * Female, 
           zRatio = (Ratio - mRatio)/sRatio)
  
  set.seed(seed)
  swim_split <- initial_split(this_swim, prop = 0.8)
  swim_train <- training(swim_split)
  swim_test <- testing(swim_split)
  
  #metric to evaluate
  metric <- metric_set(rmse) 
  ctrl_res <- control_stack_resamples()
  
  ## Must be done AFTER you standardize
  folds <- vfold_cv(swim_train, v = 5) 
  
  # Define models 
  #general recipe
 swim_rec <- 
      recipe(zRatio ~ zAge + Female + zAgeF, data = swim_train) 

  #neural net
  #model definition 
  nn_spec <- mlp(
    hidden_units = nhidden,
    penalty = 0,
    epochs = nepoch) %>%  
    set_engine("nnet") %>% 
    set_mode("regression") 
  
  #recipe extension (no extension necessary)
  nn_rec <-
    swim_rec
  
  #workflow
  nn_wflow <- 
    workflow() %>%
    add_model(nn_spec) %>%
    add_recipe(nn_rec)
  
  # fit to the 5-fold cv
  nn_res <- 
    fit_resamples(
      nn_wflow,
      resamples = folds,
      metrics = metric,
      control = ctrl_res
    )
  
  nn.res <- fit(nn_wflow, data = swim_train)
  
  #polynomial 
  #create model definition 
  lin_reg_spec <-
    linear_reg(penalty = 0.0) %>%
    set_engine("glmnet")
  
  #extend recipe
  lin_reg_rec <-
    swim_rec %>%
    step_poly(zAge, degree = ndeg)  %>%
    step_poly(zAgeF, degree = ndeg)
  
  # add both to a workflow
  lin_reg_wflow <- 
    workflow() %>%
    add_model(lin_reg_spec) %>%
    add_recipe(lin_reg_rec)
  
  # fit to the 5-fold cv
  lin_reg_res <- 
    fit_resamples(
      lin_reg_wflow,
      resamples = folds,
      metrics = metric,
      control = ctrl_res
    )
  
  lin.res <- fit(lin_reg_wflow, data = swim_train)
  
  ## Spline 
  spline_spec <-
    linear_reg() %>%
    set_engine(engine = 'lm') %>%
    set_mode('regression')
  
  spline_rec <- swim_rec %>%
    step_ns(zAge, deg_free = spdeg) %>%
    step_ns(zAgeF, deg_free = spdeg) 
  
  spline_wflow <- workflow() %>%
    add_model(spline_spec) %>%
    add_recipe(spline_rec)
  
  spline_res <- fit_resamples(
    spline_wflow,
    resamples = folds,
    metrics = metric,
    control = ctrl_res
  )
  
  spline.res <- fit(spline_wflow, data = swim_train)
  
  ## Stacked Model
  swim_st <- 
    stacks() %>%
    add_candidates(lin_reg_res) %>%
    add_candidates(nn_res) %>%
    add_candidates(spline_res) 
  
  st.res <-
    swim_st %>%
    blend_predictions() %>%
    fit_members()
  
  #keeping track of weights
  lin <- collect_parameters(st.res, "lin_reg_res")
  nn <- collect_parameters(st.res, "nn_res")
  spl <- collect_parameters(st.res, "spline_res")
  weights <- rbind(lin, nn, spl)
  
  preds <- data.frame(
    NN = predict(nn.res, new_data = swim_test)$`.pred`, 
    PL = predict(lin.res, new_data = swim_test)$`.pred`,
    SP = predict(spline.res, new_data = swim_test)$`.pred`,
    ST = predict(st.res, new_data = swim_test)$`.pred`) 
  preds$Avg <- apply(preds[,c("NN","PL", "SP")], 1, mean)
    preds$ObsTime = swim_test$zRatio
  
  ## MSE on scaled data 
  pred.rmse <- apply(preds, 2, FUN = function(y) {
    sqrt(sum((preds$ObsTime - y)^2/nrow(preds)))
  })
  
  ## Set up plotting data
  plot.dat <- expand.grid(Age = 35:84, 
                                      Female = 0:1) %>% 
    mutate(zAge = (Age - mAge) / sAge, zAgeF = Female*zAge)
  
  ## Get standard predictions 
  plot.preds <- data.frame(
    SP = predict(spline.res, new_data = plot.dat)$`.pred`, 
    NN = predict(nn.res, new_data = plot.dat)$`.pred`, 
    PL = predict(lin.res, new_data = plot.dat)$`.pred`, 
    ST = predict(st.res, new_data = plot.dat)$`.pred`)
    plot.preds <- plot.preds * sRatio + mRatio 

  plot.preds$Avg <- apply(plot.preds[, c("SP", "NN", "PL")], 1, mean)
  plot.out <- cbind(plot.dat, plot.preds)
  
  return(list(RMSE = pred.rmse, 
              PlotData = plot.out, 
              Weights = weights))
}

#one time through the function
#test <- do.one(seed = 1, swim = swimming, nhidden = 5, nepoch = 50,  ndeg = 6, spdeg = 5)
```

Run multiple iterations:

```{r}
do.one.boot <- function(seed, swim, nhidden, nepoch, ndeg, spdeg) {
  set.seed(seed)
  this.swim <- swim[sample(1:nrow(swim), replace = TRUE), ]
  res <- do.one(seed, this.swim, nhidden = nhidden, 
                nepoch = nepoch, ndeg = ndeg, spdeg = spdeg)
  return(res)
}

startseed = 1
bigB = 10

bigres.dividedres <- mclapply(1:bigB, FUN = function(i) {
  do.one.boot(startseed + i - 1, swimming,  
              nhidden = 5, nepoch = 50,  ndeg = 6, spdeg = 5)
}) 
```

Plotting the results:

```{r}
res <- bigres.dividedres

new_list<- list()
for (i in 1:bigB){
  new_list[[i]] <- res[[i]]$PlotData
}


##########
wide.preds <- do.call(rbind, new_list) %>% 
  group_by(Age, Female) %>% 
  summarize( 
    NN.mean = mean(NN), 
    NN.lo = quantile(NN, 0.025), 
    NN.hi = quantile(NN, 0.975), 
    PL.mean = mean(PL), 
    PL.lo = quantile(PL, 0.025), 
    PL.hi = quantile(PL, 0.975), 
    SP.mean = mean(SP), 
    SP.lo = quantile(SP, 0.025), 
    SP.hi = quantile(SP, 0.975), 
    ST.mean = mean(ST), 
    ST.lo = quantile(ST, 0.025), 
    ST.hi = quantile(ST, 0.975)) %>% 
  mutate(Sex = ifelse(Female == 0, "M", "F")) 

long.preds <- wide.preds %>% 
  pivot_longer(., cols = NN.mean:ST.hi, 
               names_pattern = "(.*)\\.(.*)", 
               names_to = c("model", "outval"), 
               values_to = "PredictedScore") 

plot.preds <- long.preds %>% 
  pivot_wider(names_from = outval, values_from = PredictedScore)

plot.preds <- plot.preds %>%
  mutate(Gender = ifelse(Female == "0", "M", "W"))

g <- plot.preds %>% 
  ggplot() + 
  geom_line(aes(x=Age, y=mean, group=model, col=model)) +
  facet_wrap(vars(Gender)) + 
  ggtitle("Predicted Ratio") + 
  xlab("Age (years)") + ylab("Predicted Time") + 
  theme(legend.position = "bottom") 

g

g2 <- plot.preds %>% 
  filter(model == "ST") %>% 
  ggplot() + 
  geom_line(aes(x=Age, y=mean), size=.5) + 
  geom_ribbon(aes(x=Age, ymin=lo, ymax=hi), 
              alpha=0.5, linetype=2)  + 
  facet_wrap(vars(Gender)) + 
  ggtitle("Predicted Ratio - Stacked Model", 
          sub = "Mean +/- 95% Percentile CI (1000 Bootstrap Samples)") + 
  xlab("Age (years)") + ylab("Predicted Time") + 
  theme(legend.position = "bottom") 

g2
```



